---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# neonWaterQuality

<!-- badges: start -->
<!-- badges: end -->

This package provides high-frequency water quality data from three NEON sites (ARIK, CARI, LEWI) for 2018-2019, as analyzed in Kermorvant et al. (2023). It also includes a Shiny application to explore the data interactively.

This package was created for Assessment 4 of ETC5523 (Communicating with Data).

## Installation

You can install the development version of neonWaterQuality from [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-sagabmonash) with:

``` {r}
# install.packages("remotes")
remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-sagabmonash")
```

## Quick Start

To explore the data, you can launch the interactive Shiny app:

```{r}
library(neonWaterQuality)
launch_app()
```

## Website

The full documentation and package website is available at: [GitHub](https://github.com/ETC5523-2025/assignment-4-packages-and-shiny-apps-sagabmonash)